Indicators for Ecosystem Condition in Norway

### 2023-12-14 birdIndicators.Rmd

## Introduction {#intro-bb}
The Norwegian common breeding bird monitoring scheme ("hekkefuglovervåking" in Norwegian; formerly called TOV-E) was established between 2005 and 2010. The main goal of the programme is to gather data on the abundance of the most common terrestrial bird species to assess population trends and changes across Norway (available at the website), and help proposing efficient management and conservation measures for these species and their habitats in Norway. The "hekkefuglovervåking" is coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), and has replaced predecessor monitoring schemes that ran from 1996 to 2008.

These data collected through "hekkefuglovervåkingen" form the basis for a variety of national and international assessments including nationally reported species population trends (https://hekkefuglovervakingen.nina.no/hekkefugl), updates to the Red List of Birds in Norway following the IUCN protocol (https://artsdatabanken.no/lister/rodlisteforarter), indicators for the Norwegian Nature Index (https://www.naturindeks.no/), and single- and multi-species indicators that are reported nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/) and internationally to e.g. the Pan European Bird Monitoring Scheme (PECBMS, https://pecbms.info/) and the OECD.

Workflows generating outputs and numbers for several of these reporting routes are already integrated thanks to coordination and standardization efforts by PECMBS, and the calculation of breeding bird indicators for use in ecosystem condition assessments should be incorporated into the same workflow. We therefore suggest to use the same multi-species indicators in the framework for ecosystem condition assessment as are reported nationally and internationally: the established "Farmland bird index", "Mountain bird index", and "Forest bird index", as well as a corresponding index for wetland birds.

## About the underlying data {#underlying-bb}
The data from "hekkefuglovervåkingen" are gathered mostly by experienced volunteer bird-watchers following a standardized sampling protocol (https://hekkefuglovervakingen.nina.no/Fugl/public/papirskjema/MethodologyEng.pdf). Observers will visit each route once a year between 20. May and 10. June (depending on latitude) and carry out the census in good weather conditions. The observers record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route. A total of 229 species are typically recorded during this census. After 5 minutes, the observer walks to the next point in a straight line (when possible). During this 'transect' the observer is requested to record also a subset of species that are considered rare or difficult to capture during the 5-minute point counts (n = 121).

@@

### Representativity in time and space {#rep-in-time-and-space-bb}
"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats present in Norway. Each route has between 12 and 20 counting points (ca. 9000 across Norway), which typically are 300 m apart, describing a square with side 1.5 km long (the observer walks on a straight line between two consecutive points). Deviations of this typical route shape are normally due to the landscape configuration (e.g. rough terrain, presence of a lake or river, sea, glacier, etc.). Due to limited resources, only a random sample of 370 routes are surveyed every year with an aim to survey at least 80% of the routes yearly.

This monitoring scheme is therefore designed to gather representative samples in all (common) habitats along both latitudinal and altitudinal gradients.

Even though implementation of the monitoring scheme started in 2005, full coverage at the national scale was not achieved until 2010.

### Original units {#original-units-bb}
The observers in the field record "number of individuals" seen or heard during the census. Those values are then converted into "number of pairs" by multiplying the number of individuals by a conversion factor, which is 2 or 5 depending on the species (for national reporting only).
In the first step of the workflow, the raw counts are converted into species-specific abundance indices, defined as % of the population in a given baseline year (in Norway this is typically 2008). The abundance indices then form the basis of further analyses and aggregation routines.

### Temporal coverage {#temp-coverage-bb}
The "hekkefuglovervåkingen" was established in 2005 in central Norway and expanded from there to reach full national coverage by 2010. For many species, data is already of good coverage and high quality starting in 2008. The monitoring programme is still ongoing as of 2023.

Predecessor monitoring schemes existed prior to the establishment of "hekkefuglovervåkingen" and led to the collection of decent data in the period from 1996 up to and including 2008. 
Although the predecessor monitoring programmes and "hekkefuglovervåkingen" are not the same, the overlap for the years 2007 and 2008 makes it possible to combine abundance indices from across time using tools like "RSWAN", which is provided by PECBMS. For a subset of 55 species, longer time-series of abundance indices (1996-2022, as of now) are therefore available.

### Aditional comments about the dataset {#additional-comments-bb}
From all species recorded within the Norwegian common breeding bird monitoring scheme (n = 229), 85 have sufficient data for abundance trend analyses. Abundance indices of 71 of those species, representing those with wide distributions and high continental-scale relevance, are reported annually to PECBMS. Longer (combined) time series can be calculated for 55 species, and 54 of those serve as indicators in the Norwegian Nature Index. The established breeding bird indicators contain varying numbers of species: the farmland bird index has 7 species, the mountain bird index has 8 species, and the forest bird index has 24 species. For the new index for wetland birds, we used 7 out of the 8 species suggested for this purpose by @husby2011. The one species we dropped was the Common Crane _(Grus grus)_ as it has seen strong recent population increases that are linked to anthropogenic impacts unrelated to the status of their wetland breeding habitats (Bård Stokke, pers. communication).

@@

### Data sets {#datasets-bb}
The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåkingen" and the calculated abundance indices from the predecessor monitoring schemes. Additionally, we also use a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow.

#### Monitoring data from "Hekkefuglovervåkingen"
The route-based raw data from "hekkefuglovervåkingen" are compiled after each breeding season and stored in NINA's database. From that database we download the data relevant for the PECBMS Trim analyses and for the specified range of years (here from 2006 up to and including 2022). The function ``downloadData_TRIM`` does that for a specified year range (from ``minYear`` up to and including ``maxYear``), and requires specifying which driver is used for accessing the database (via argument ``DriverName``). When running from the NINA RStudio server, we need to use "FreeTDS". When running locally from NINA computers, we need to select "SQL Server".
Note that downloading this data requires user-specific access privileges which can be requested from NINA.

```{r, eval = FALSE}
minYear <- 2006
maxYear <- 2022

@@

#### PECBMS Trim results from predecessor monitoring programme
The abundance indices generated by the PECBMS RTRIM workflow using data from the predecessor scheme are stored in NINAs local file repositories. Unlike the Trim analyses for the data from "hekkefuglovervåkingen", which are re-run every year after new data has been added, the old abundance indices remain unchanged as no new data are being added.
These data are needed for the RSWAN workflow, and they need to be present in the respective "working_folder". So we copy them from their location in NINA's local file storage into the correct folder for analyses.

@@

The table ``Spp_selection`` contains information for each species on which reporting routes is should be included in. The multispecies indices are grouped into a single category ("MSI") here, but we can extract the species that go into the MSI for each ecosystem via the additionally provided lists.

@@

As per now, the breeding bird indicators are calculated at the national scale only and the data is therefore not linked to any geofiles here.

@@

In the next step, the downloaded raw data get analysed and converted into species-specific abundance indices using the RTRIM approach ("**TR**ends and **I**ndices for **M**onitoring data", @R-rtrim2020). The RTRIM analyses are run for a subset of 85 species which have sufficient data for it and using the standardized scripts provided by PECBMS (more information on https://pecbms.info/methods/software/trim/). 
For the application here, we have written wrapper function around the PECBMS code for each step of the RTRIM workflow.

@@

The results of the the RTRIM workflow are a series of .csv and .txt files. These are delivered directly to PECBMS each year after the analyses have been run. Prior to delivery, they get copied into a designated folder:

@@

For 55 of the 85 species, RTRIM analyses have also been run using data from the predecessor monitoring schemes. The RTRIM output files from both old and new analyses should now be present in the same folder. 
From there, they enter into the next step of the workflow: PECBMS' RSWAN routine. This is based on another set of scripts distributed through PECBMS which combine RTRIM outputs from different monitoring schemes / time periods.

@@

THe RSWAN routine produces output files for the selected 55 species that are equivalent to those produced by the RTRIM workflow, but have a prefix "COMB" to indicate that they are based on combined time series.

@@

The breeding bird indicators for ecosystem condition assessments are mutli-species indices (MSIs) calculated following the guidelines from PECBMS (https://www.cbs.nl/en-gb/society/nature-and-environment/indices-and-trends--trim--/msi-tool) and using the algorithm of @soldaat2017.

@@

The function also requires us to provide a name (argument) and a list of EURING codes of species to include (argument ). We list these for all four ecosystem indices:

@@

  # Combine and add additional columns
  MSI_export[[i]] <- cbind(mean_wide, sd_wide) %>%
    dplyr::mutate(reference_high = NA, 
                  reference_low = 0, 
                  thr = NA)
}
```
Notably, I have not added the ``i_YYYY` columns yet as these can only be calculated once we have discussed and come to an agreement on reference value (year), threshold value, and scaling function.

>>>>

## Introduction {#intro-bb}
The Norwegian common breeding bird monitoring scheme ("hekkefuglovervåking" in Norwegian; formerly called "TOV-E") was established between 2005 and 2010. The main goal of the scheme is to gather data on the abundance of the most common terrestrial bird species to assess population trends and changes across Norway (results can be found at the website), and help proposing efficient management and conservation measures for these species and their habitats in Norway. The "hekkefuglovervåking" is coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), and replaced a predecessor monitoring scheme that ran from 1996 to 2008.

The data collected through "hekkefuglovervåkingen" form the basis for a variety of national and international assessments including nationally reported species population trends (https://hekkefuglovervakingen.nina.no/hekkefugl), updates to the Red List of Birds in Norway following the IUCN protocol (https://artsdatabanken.no/lister/rodlisteforarter), indicators for the Norwegian Nature Index (https://www.naturindeks.no/), and single- and multi-species indicators that are reported nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/) and internationally to e.g. the Pan European Bird Monitoring Scheme (PECBMS, https://pecbms.info/) and the OECD.

Workflows generating outputs and numbers from all these reporting routes are already integrated thanks to coordination and standardization efforts by PECMBS, and the calculation of breeding bird indicators for use in ecosystem condition assessments should be incorporated into the same workflow. We therefore suggest to use the same multi-species indicators in the framework for ecosystem condition assessment as are reported nationally and internationally: the established "Farmland bird index", "Mountain bird index", and "Forest bird index", as well as a corresponding index for wetland birds.

## About the underlying data {#underlying-bb}
The data from "hekkefuglovervåkingen" are gathered mostly by experienced volunteer bird-watchers following a standardized sampling protocol (https://hekkefuglovervakingen.nina.no/Fugl/public/papirskjema/MethodologyEng.pdf). Observers will visit each route once a year between 20. May and 10. June (depending on latitude) and carry out the census in good weather conditions. The observers record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route (see below for further information on the sampling design). A total of 229 species are typically recorded during this census. After 5 minutes, the observer walks to the next point in a straight line (when possible). During this 'transect' the observer is requested to record also a subset of species that are considered rare or difficult to capture during the 5-minute point counts (n = 121).

@@

### Representativity in time and space {#rep-in-time-and-space-bb}
"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats along both latitudinal and altitudinal gradients present in Norway. Each route has between 12 and 20 counting points (ca. 9000 across Norway), which typically are 300 m apart, describing a square with side 1.5 km long (the observer walks on a straight line between two consecutive points). Deviations of this typical route shape are normally due to the landscape configuration (e.g. rough terrain, presence of a lake or river, sea, glacier, etc.). Due to limited resources, only a random sample of 370 routes are surveyed every year with an aim to survey at least 80% of the routes yearly.

Even though implementation of the monitoring scheme started in 2005, full geographical coverage at the national scale was not achieved until 2010.

### Original units {#original-units-bb}
The observers in the field record "number of individuals" seen or heard during the census. Those values are then converted into "number of pairs" by multiplying the number of individuals by a conversion factor, which is 2 or 5 depending on the species (for national reporting only).
In the first step of the workflow, the raw counts are converted into species-specific abundance indices, defined as % of change relative to the population in a given baseline year (in Norway this is typically 2008). The abundance indices then form the basis of further analyses and aggregation routines.

### Temporal coverage {#temp-coverage-bb}
The "hekkefuglovervåkingen" was established in 2005 in central Norway and expanded from there to reach full national coverage by 2010. For many species, data are already of good coverage and high quality starting in 2008. The monitoring scheme is still ongoing as of 2023.

A predecessor monitoring scheme existed prior to the establishment of "hekkefuglovervåkingen" and led to the collection of decent data in the period from 1996 up to and including 2008. 
Although the predecessor monitoring programmes and "hekkefuglovervåkingen" are not the same, the overlap for the years 2007 and 2008 makes it possible to combine abundance indices from across time using tools like "RSWAN", which is provided by PECBMS. For a subset of 55 species, longer time-series of abundance indices (1996-2023, as of now) are therefore available.

### Aditional comments about the dataset {#additional-comments-bb}
NINA perfoems quality-checks of the data delivered by the observers to ensure that the new data entered to the database are reliable. In addition, NINA completes the annual detaset by entering the zeroes (undetected species) at surveyed routes and -1s (for not surveyed routes). This ensures that all species have data (either a positive count, a zero or a -1) for each route and year.

From all species recorded within the Norwegian common breeding bird monitoring scheme (n = 229), 85 have sufficient data for abundance trend analyses. Abundance indices of 71 of those species, representing those with wide distributions and high continental-scale relevance, are reported annually to PECBMS. Longer (combined) time series can be calculated for 55 species, and 54 of those serve as indicators in the Norwegian Nature Index. The established breeding bird indicators contain varying numbers of species: the farmland bird index has 7 species, the mountain bird index has 8 species, and the forest bird index has 24 species. For the new index for wetland birds, we used 7 out of the 8 species suggested for this purpose by @husby2011. The one species we dropped was the Common Crane _(Grus grus)_ as it has seen strong recent population increases that are linked to anthropogenic impacts unrelated to the status of their wetland breeding habitats (Bård Stokke, pers. communication).

@@

### Data sets {#datasets-bb}
The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåking" and the calculated abundance indices from the predecessor monitoring scheme. Additionally, we also use a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow as the different outcomes require different information.

#### Monitoring data from "Hekkefuglovervåking"
The route-based raw data from "hekkefuglovervåking" are compiled after each breeding season and curated and stored in NINA's database. Before storing new data in the database, a thorough quality-check is performed. From that database we download the data relevant for the PECBMS Trim analyses and for the specified range of years (here from 2006 up to and including 2023). The function ``downloadData_TRIM`` does that for a specified year range (from ``minYear`` up to and including ``maxYear``), and requires specifying which driver is used for accessing the database (via argument ``DriverName``). When running from the NINA RStudio server, we need to use "FreeTDS". When running locally from NINA computers, we need to select "SQL Server".
Note that downloading these data requires user-specific access privileges which can be requested from NINA IT department ('datahjelp').

```{r, eval = FALSE}
minYear <- 2006
maxYear <- 2023

@@

#### PECBMS Trim results from predecessor monitoring programme
The abundance indices generated by the PECBMS RTRIM workflow using data from the predecessor scheme are stored in NINA's local file repositories. Unlike the Trim analyses for the data from "hekkefuglovervåking", which are re-run every year after new data has been added, the old abundance indices remain unchanged as no new data are being collected.
These data are needed for the RSWAN workflow, and they need to be present in the respective "working_folder". Hence, we copy the files from their location in NINA's local file storage into the correct folder for analyses.

@@

The table ``Spp_selection`` contains information for each species on which reporting routes should be included in. The multispecies indices are grouped into a single category ("MSI") here, but we can extract the species that go into the MSI for each ecosystem via the additionally provided lists.

@@

As per now, the breeding bird indicators are calculated at the national scale only and the data are therefore not linked to any geofiles here.

@@

In the next step, the downloaded raw data is analysed and converted into species-specific abundance indices using the RTRIM approach ("**TR**ends and **I**ndices for **M**onitoring data", @R-rtrim2020). The RTRIM analyses are run for a subset of 85 species which have sufficient data for it using the standardized scripts provided by PECBMS (more information on https://pecbms.info/methods/software/trim/). 
For the application here, we have written wrapper function around the PECBMS code for each step of the RTRIM workflow.

@@

The results of the the RTRIM workflow are a series of .csv and .txt files. These are delivered directly to PECBMS each year after the analyses have been run. Prior to delivery, they are copied into a designated folder:

@@

For 55 of the 85 species with sufficient data, RTRIM analyses have also been run using data from the predecessor monitoring scheme. The RTRIM output files from both old and new analyses should now be present in the same folder. 
From there, they enter into the next step of the workflow: PECBMS' RSWAN routine. This is based on another set of scripts distributed through PECBMS which combine RTRIM outputs from different monitoring schemes / time periods.

@@

The RSWAN routine produces output files for the selected 55 species that are equivalent to those produced by the RTRIM workflow, but have a prefix "COMB" to indicate that they are based on combined time series.

@@

The breeding bird indicators for ecosystem condition assessments are mutlispecies indices (MSIs) calculated following the guidelines from PECBMS (https://www.cbs.nl/en-gb/society/nature-and-environment/indices-and-trends--trim--/msi-tool) and using the algorithm of @soldaat2017.

@@

The function also requires us to provide a name (argument) and a list of EURING codes of species to include (argument). We list these for all four ecosystem indices:

@@

  # Combine and add additional columns
  MSI_export[[i]] <- cbind(mean_wide, sd_wide) %>%
    dplyr::mutate(reference_high = NA, 
                  reference_low = 0, 
                  thr = NA)
}
```

Notably, I have not added the ``i_YYYY` columns yet as these can only be calculated once we have discussed and come to an agreement on reference value (year), threshold value, and scaling function.
